Creep and brittle failure of a protein gel under stress

### 2014-01-31 creep/creep_yaourt.tex
Intro: Single equation, fundamental issues remain unexplored no slippage at the Plexiglas walls Monkman-Grant relation no self healing Re-work conclusion SI: sigma and diameter of plane-plane

\begin{abstract}
Biomaterials such as protein or polysaccharide gels are known to behave qualitatively as soft solids in that they rupture under an external load. Combining optical and ultrasonic imaging to shear rheology we show that the failure scenario of a protein gel is typical of brittle solids: after a power-law creep regime fully accounted for by linear viscoelasticity and homogeneous deformation, fractures nucleate and grow logarithmically perpendicularly to shear up to sudden rupture. The failure time surprisingly follows a decreasing power-law with the applied shear stress strongly reminiscent of the Basquin law of fatigue for solids. These results are in excellent agreement with recent fiber-bundle models that include damage accumulation on elastic fibers and exemplify protein gels as model brittle soft solids.
\end{abstract}

@@

Biogels formed through the self-association of polysaccharide coils, collagen, actin filaments or attractive globular proteins play a major role in biochemistry and microbiology \cite{Viovy:2000}, biological networks and cell mechanics \cite{Stricker:2010} as well as in food science \cite{Mezzenga:2005}. These biomaterials all behave as elastic solids under small deformations but display remarkable nonlinear behavior generally featuring stress- or strain-stiffening \cite{Gardel:2004} and fractures prior to irreversible rupture \cite{Bonn:1998,Baumberger:2006,Daniels:2007}. Irreversibility stems from the existence of an external control parameter, e.g. temperature or pH in the case of thermoreversible or acid-induced gels respectively, which makes such biogels impossible to shear-rejuvenate after failure, contrary to other soft glassy materials such as emulsions, colloidal gels and glasses \cite{Cloitre:2000,Divoux:2012,Siebenburger:2012}. So far huge effort has been devoted to designing protein gels with specific properties and textures at rest \cite{Dickinson:2006,Gibaud:2012}. However, their mechanical behavior under a shear stress deep into the nonlinear regime has only been partially addressed \cite{Higgs:1990,vanVliet:1995} and several simple issues remain unexplored such as the spatially resolved rupture scenario or the physical relevance of the analogy brittle failure in hard solids.

@@

Gels are prepared by dissolving sodium caseinate powder (Firmenich) at 4\% wt. in deionized water under gentle mixing at 35$^{\circ}$C and 500~rpm. To induce gelation, 1\% wt. glucono-$\delta$-lactone (GDL) in powder (Firmenich) is dissolved in the solution and its hydrolysis progressively lowers the pH over the course of 8~hours [Fig.~\ref{suppfig1}(a) in the Supplemental Material]. While still liquid, the solution is poured into the gap of a polished Plexiglas Couette cell immersed into a temperature-controlled water tank at 25.0$\pm$0.1$^{\circ}$C \cite{note}. Rheological data are recorded during gel formation by a stress-controlled rheometer through small amplitude oscillatory shear at frequency $f=1$~Hz (Fig.~1 in the Supplemental Material) \cite{note}. Gelation is complete when the elastic ($G'$) and viscous ($G''$) moduli reach a plateau with $G'\gg G''$. A constant stress $\sigma$ is then applied to the sample at time $t=0$ and the subsequent strain response $\gamma(t)$ is monitored. Images of the gel are recorded simultaneously to the rheology (Logitech Webcam Pro 9000). The local velocity and strain fields can also be imaged in the gradient--vorticity plane $(r,z)$ simultaneously to rheology by a custom-made ultrasonic scanner detailed in \cite{Gallot:2013}. In this case, prior to acidification, the sodium caseinate solution is seeded with acoustic tracers here 3\% wt. polyamide spheres (Orgasol 2002 ES3 NAT 3, Arkema, diameter 30~$\mu$m, density 1.02) that do not modify the final gel properties [Fig.~\ref{suppfig1}(b) in the Supplemental Material]. Failure being irreversible, each creep experiment requires to prepare a fresh sample in-situ.

@@

As seen in the inset of Fig.~\ref{fig1}, the failure time $\tau_f$ sharply decreases as a power-law of $\sigma$ with an exponent $\beta\simeq 5.5$. $\tau_f$ further allows us to rescale all the shear rate data $\gp(t)$ onto the single master curve of Fig.~\ref{fig2}(a) by plotting $\gp/\gp_{\rm min}$ vs $t/\tau_f$, where $\gp_{\rm min}$ is the minimum shear rate reached at a time $\tau_{\rm min}$ [see also Fig.~\ref{suppfig2} in the Supplemental Material for the original $\gp(t)$ data]. On about four decades for $t\lesssim 0.1\tau_f$, the shear rate decreases as a power-law $\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$. This is strongly reminiscent of the {\it primary creep} observed in solids and referred to as Andrade creep \cite{Andrade:1910,Miguel:2002,Nechad:2005}. Interestingly, here, this scaling can be fully accounted for by linear viscoelasticity without invoking any irreversible plasticity. Indeed casein gels display a power-law rheology $G'(f)\sim G''(f)\sim f^{0.15}$ [inset of Fig.~\ref{fig2}(a)], which corresponds to a compliance $J(t)\equiv\gamma(t)/\sigma\sim t^{0.15}$ in the linear deformation regime \cite{Jaishankar:2013}. This nicely corresponds to the observed $\gp(t)\sim t^{-0.85}$ and suggests that primary creep is purely reversible without unrecoverable plastic events. Homogeneity and linear response are further confirmed, at least up to the available spatial ($\sim 10~\mu$m) and temporal ($\sim 1$~s) resolutions, by direct visualization which shows no sign of crack or fracture [Fig.~\ref{fig3}(a)] and by ultrasonic imaging which reveals velocity and strain fields averaged over the vertical direction $z$ that linearly decrease with the radial position $r$ within the gap [Fig.~\ref{fig4}(a-b)] with insignificant $z$-dependence [Fig.~\ref{fig3}(a, left)].

For $0.1\lesssim t/\tau_f\lesssim 0.9$, $\gp(t)$ departs from power-law behavior and goes through a minimum value at time $\tau_{\rm min}= (0.56\pm0.04)\tau_f$ independently of the applied stress [Fig.~\ref{fig2}(b) and inset] as similarly reported for solid composite materials \cite{Nechad:2005} and fiber-bundles models (FBMs) \cite{Kovacs:2008,Jagla:2011}. During this {\it secondary creep} regime, regularly-spaced cracks nucleate from the top and bottom edges of the Couette cell and start growing perpendicular to the applied stress [Fig.~\ref{fig3}(b), see also Movie~2 in the Supplemental Material]. These macroscopic fractures are invaded with water expelled from the surrounding gel matrix. Although fractures have not yet entered the ultrasonic region of interest, velocity maps become heterogeneous along the $z$ direction [Fig.~\ref{fig3}(b, left)] and display intermittent fluctuations [Fig~\ref{fig4}(c,d)]. The level of these fluctuations, which can also be seen on the global response for the same applied stress [$\blacktriangledown$ in Fig.~\ref{fig2}(b)], is poorly reproducible and stress-dependent. We propose that such intermittency results from self-healing microscopic cracks which precede the tip of the fracture. Since heterogeneity is much smaller along the radial direction [compare red and black lines in Fig~\ref{fig4}(d)], these ``precursors'' either extend across the full 2-mm gap or induce long-range displacements that span the gap. In any case, such noisy features are typical of slow crack growth in solids \cite{Vanel:2009,Kun:2007} and more generally of avalanche-like behavior \cite{Sethna:2001}.

@@

Let us now summarize and discuss the most prominent results of this Letter. First we have shown that casein gels display a remarkable failure scenario similar to that of brittle solids and characterized by the same three successive creep regimes. Here, the primary creep and its power-law exponent [$\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$] are fully accounted for by linear viscoelasticity and local measurements consistently display a homogeneous bulk deformation. Such a link between creep and viscoelasticity is shared by other biopolymer gels with power-law rheology \cite{Jaishankar:2013,Gobeaux:2010} but contrasts with the primary creep of solid materials where smaller exponents are found (0.3 $\lesssim \alpha \lesssim 0.7$). Besides primary creep in solids is irreversible and has been attributed to dislocation interactions in the case of crystalline materials \cite{Miguel:2002} or to localized plastic rearrangements in amorphous materials \cite{Rosti:2010}. Protein gels rather appear to remain purely elastic and reversible throughout primary creep, a hallmark of brittle materials.

The logarithmic fracture growth in the tertiary creep regime constitutes our second important result. Such an evolution is also commonly reported in disordered solid materials displaying brittle rupture (such as paper) and interpreted in the framework of Griffith-like models based on global or local energy barriers \cite{Vanel:2009}. However these approaches all predict exponential scalings for $\tau_f(\sigma)$ while our data are best fitted by the decreasing power-law $\tau_f \sim \sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ [inset of Fig.~\ref{fig1}]. This last key result suggests that thermally activated crack growth is not relevant to the present soft system. Rather the power-law scaling is strikingly reminiscent of the Basquin law of fatigue found for a variety of heterogeneous or cellular materials under cyclic deformation \cite{Kun:2007,Basquin:1910,Suresh:2006}. Basquin law has also been recently predicted by FBMs that combine elastic fibers with a local yield strain and take into account damage accumulation \cite{Kun:2007,Halasz:2012}. Interestingly assuming damage accumulation to be proportional to $\sigma^\gamma$ directly leads to Basquin law with $\beta=\gamma$ and large values of $\gamma$, typically larger than 5 as found here for $\beta$, lead to macroscopic cracks due to the simultaneous rupture of a large number of fibers \cite{Halasz:2012}. More generally FBMs under compressive or elongational load predict three successive creep regimes exactly alike Fig.~\ref{fig2} with similar proportionality between $\tau_{\rm min}$ and $\tau_f$ and finite-time singularity \cite{Nechad:2005,Jagla:2011}. This highlights the relevance of FBMs in the context of creep in protein gels, whose microstructure indeed appears to be formed of strands \cite{Kalab:1983,Roefs:1990}, but also urges to study FBMs in shear geometries to check whether they would be able to predict fracture growth as observed here.

Finally, we emphasize that the (reversible) Andrade-like creep and the (irreversible) crack growth are two  physical processes that effectively superimpose to yield the global rheological response. Indeed, as seen from the yellow line in Fig.~\ref{fig2}, the master curve $\gp(t)/\gp_{\rm min}$ vs $t/\tau_f$ is perfectly fitted by:
%the master curve $\tilde{\gp}(t)=\gp(t)/\gp_{\rm min}$ vs $\tilde{t}=t/\tau_f$ is perfectly fitted by:
\begin{equation}
\frac{\gp(t)}{\gp_{\rm min}}=\lambda \left(\frac{t}{\tau_f}\right)^{-\alpha}+ \frac{\mu}{1-t/\tau_f}\,,
%\tilde{\gp}(\tilde{t})=\lambda \tilde{t}^{-\alpha}+ \frac{\mu}{1-\tilde{t}}\,,
\label{e.fit}
\end{equation}
with only two adjustable parameters $\lambda$ and $\mu$ once $\alpha=0.85$ is fixed. This allows us to interpret the secondary creep regime as a mere crossover from creep to crack growth. To conclude, the present time- and space-resolved study shows that protein gels qualify as model, brittle soft solids. The next step consists in a statistical study of the fluctuations associated to crack nucleation and growth as well as a systematic investigation of other systems in order to check for universality in the irreversible creep rupture of soft solids in general and of biogels in particular. Such a study is expected to have important implications in understanding the behavior of biomaterials under extreme stress conditions.

@@

\bibitem{Mezzenga:2005} R. Mezzenga, P. Schurtenberger, A. Burbidge and M. Michel, Nature Materials  {\bf 4}, 729--740 (2005).

@@

\bibitem{Baumberger:2006} T. Baumberger, C. Caroli and D. Martina, Nature Materials {\bf 5}, 552-555 (2006).

@@

\bibitem{Higgs:1990} P.G. Higgs and S.B. Ross-Murphy, Int. J. Biol. Macromol. {\bf 12}, 233--240 (1990).

@@

@@

@@

Supplemental Movie~1 shows the failure of a 4\%~wt. casein gel acidified with 1\%~wt. GDL in a plate-plate geometry for an imposed shear stress $\sigma=XXX$~Pa. The plate diameter is 40XXX~mm and the gap width is 1XXX~mm. Fractures grow parallel to the vorticity direction, i.e. along the radial direction in this case. The global rheological response is fully similar to that measured in the Couette geometry.

>>>>

\begin{abstract}
Biomaterials such as protein or polysaccharide gels are known to behave qualitatively as soft solids in that they rupture under an external load. Combining optical and ultrasonic imaging to shear rheology we show that the failure scenario of a protein gel is typical of brittle solids: after a power-law creep regime fully accounted for by linear viscoelasticity and homogeneous deformation, fractures nucleate and grow logarithmically perpendicularly to shear up to sudden rupture. A single equation accounting for those two successive processes nicely captures the full rheological response. The failure time surprisingly follows a decreasing power-law with the applied shear stress strongly reminiscent of the Basquin law of fatigue for solids. These results are in excellent agreement with recent fiber-bundle models that include damage accumulation on elastic fibers and exemplify protein gels as model brittle soft solids.
\end{abstract}

@@

Biogels formed through the self-association of polysaccharide coils, collagen, actin filaments or attractive globular proteins play a major role in biochemistry and microbiology \cite{Viovy:2000}, biological networks and cell mechanics \cite{Stricker:2010} as well as in food science \cite{Mezzenga:2005}. These biomaterials all behave as elastic solids under small deformations but display remarkable nonlinear behavior generally featuring stress- or strain-stiffening \cite{Gardel:2004} and fractures prior to irreversible rupture \cite{Bonn:1998,Baumberger:2006,Daniels:2007}. Irreversibility stems from the existence of an external control parameter, e.g. temperature or pH in the case of thermoreversible or acid-induced gels respectively, which makes such biogels impossible to shear-rejuvenate after failure, contrary to other soft glassy materials such as emulsions, colloidal gels and glasses \cite{Cloitre:2000,Divoux:2012,Siebenburger:2012}. So far huge effort has been devoted to designing protein gels with specific properties and textures at rest \cite{Dickinson:2006,Gibaud:2012}. However, their mechanical behavior under a shear stress deep into the nonlinear regime has only been partially addressed \cite{vanVliet:1995} and several fundamental issues remain unexplored such as the spatially resolved rupture scenario or the physical relevance of the analogy brittle failure in hard solids.

@@

Gels are prepared by dissolving sodium caseinate powder (Firmenich) at 4\% wt. in deionized water under gentle mixing at 35$^{\circ}$C and 500~rpm. To induce gelation, 1\% wt. glucono-$\delta$-lactone (GDL) in powder (Firmenich) is dissolved in the solution and its hydrolysis progressively lowers the pH over the course of 8~hours [Fig.~\ref{suppfig1}(a) in the Supplemental Material]. While still liquid, the solution is poured into the gap of a polished Plexiglas Couette cell immersed into a temperature-controlled water tank at 25.0$\pm$0.1$^{\circ}$C \cite{note}. Rheological data are recorded during gel formation by a stress-controlled rheometer through small amplitude oscillatory shear at frequency $f=1$~Hz (Fig.~1 in the Supplemental Material) \cite{note}. Gelation is complete when the elastic ($G'$) and viscous ($G''$) moduli reach a plateau with $G'\gg G''$. A constant stress $\sigma$ is then applied to the sample from time $t=0$ and the subsequent strain response $\gamma(t)$ is monitored. Images of the gel are recorded simultaneously to the rheology (Logitech Webcam Pro 9000). The local velocity and strain fields can also be imaged in the gradient--vorticity plane $(r,z)$ simultaneously to rheology by a custom-made ultrasonic scanner detailed in \cite{Gallot:2013}. In this case, prior to acidification, the sodium caseinate solution is seeded with acoustic tracers here 3\% wt. polyamide spheres (Orgasol 2002 ES3 NAT 3, Arkema, diameter 30~$\mu$m, density 1.02) that do not modify the final gel properties [Fig.~\ref{suppfig1}(b) in the Supplemental Material]. Failure being irreversible, each creep experiment requires to prepare a fresh sample in-situ.

@@

As seen in the inset of Fig.~\ref{fig1}, the failure time $\tau_f$ sharply decreases as a power-law of $\sigma$ with an exponent $\beta\simeq 5.5$. $\tau_f$ further allows us to rescale all the shear rate data $\gp(t)$ onto the single master curve of Fig.~\ref{fig2}(a) by plotting $\gp/\gp_{\rm min}$ vs $t/\tau_f$, where $\gp_{\rm min}$ is the minimum shear rate reached at a time $\tau_{\rm min}$ [see also Fig.~\ref{suppfig2} in the Supplemental Material for the original $\gp(t)$ data]. On about four decades for $t\lesssim 0.1\tau_f$, the shear rate decreases as a power-law $\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$. This is strongly reminiscent of the {\it primary creep} observed in solids and referred to as Andrade creep \cite{Andrade:1910,Miguel:2002,Nechad:2005}. Interestingly, here, this scaling can be fully accounted for by linear viscoelasticity without invoking any irreversible plasticity. Indeed casein gels display a power-law rheology $G'(f)\sim G''(f)\sim f^{0.15}$ [inset of Fig.~\ref{fig2}(a)], which corresponds to a compliance $J(t)\equiv\gamma(t)/\sigma\sim t^{0.15}$ in the linear deformation regime \cite{Tschoegl:1989}. This nicely corresponds to the observed $\gp(t)\sim t^{-0.85}$ and suggests that primary creep is purely reversible without unrecoverable plastic events. Homogeneity and linear response are further confirmed, at least up to the available spatial ($\sim 10~\mu$m) and temporal ($\sim 1$~s) resolutions, by direct visualization which shows no sign of crack or fracture [Fig.~\ref{fig3}(a)] and by ultrasonic imaging which reveals velocity and strain fields averaged over the vertical direction $z$ that linearly decrease with the radial position $r$ within the gap [Fig.~\ref{fig4}(a-b)] with insignificant $z$-dependence [Fig.~\ref{fig3}(a, left)] and no slippage at the Plexiglas walls.

For $0.1\lesssim t/\tau_f\lesssim 0.9$, $\gp(t)$ departs from power-law behavior and goes through a minimum value at time $\tau_{\rm min}= (0.56\pm0.04)\tau_f$ independently of the applied stress [Fig.~\ref{fig2}(b) and inset] as similarly reported for solid composite materials \cite{Nechad:2005} and fiber-bundles models (FBMs) \cite{Kovacs:2008,Jagla:2011}. This linearity between $\tau_{\rm min}$ and $\tau_f$ is also known as the Monkman-Grant relation in the context of metals \cite{Sundararajan:1989} and allows one to ``predict'' the failure time from the intermediate-time response. During this {\it secondary creep} regime, regularly-spaced cracks nucleate from the top and bottom edges of the Couette cell and start growing perpendicular to the applied stress [Fig.~\ref{fig3}(b), see also Movie~2 in the Supplemental Material]. These macroscopic fractures are invaded with water expelled from the surrounding gel matrix. Although fractures have not yet entered the ultrasonic region of interest, velocity maps become heterogeneous along the $z$ direction [Fig.~\ref{fig3}(b, left)] and display intermittent fluctuations [Fig~\ref{fig4}(c,d)]. The level of these fluctuations, which can also be seen on the global response for the same applied stress [\textcolor{red!50!black}{$\blacktriangledown$} in Fig.~\ref{fig2}(b)], is poorly reproducible and stress-dependent. We propose that such intermittency arises from microscopic cracks ahead of the fracture tip. Since heterogeneity is much smaller along the radial direction [compare red and black lines in Fig~\ref{fig4}(d)], these ``precursors'' either extend across the full 2-mm gap or induce long-range displacements that span the gap. In any case, such noisy features are typical of slow crack growth in solids \cite{Vanel:2009,Kun:2007} and more generally of avalanche-like behavior \cite{Sethna:2001}.

@@

Finally, we emphasize that the (reversible) Andrade-like creep and the (irreversible) crack growth are two  physical processes that effectively superimpose to yield the global rheological response. Indeed, as seen from the yellow line in Fig.~\ref{fig2}, the master curve $\gp(t)/\gp_{\rm min}$ vs $t/\tau_f$ is perfectly fitted by:
%the master curve $\tilde{\gp}(t)=\gp(t)/\gp_{\rm min}$ vs $\tilde{t}=t/\tau_f$ is perfectly fitted by:
\begin{equation}
\frac{\gp(t)}{\gp_{\rm min}}=\lambda \left(\frac{t}{\tau_f}\right)^{-\alpha}+ \frac{\mu}{1-t/\tau_f}\,,
%\tilde{\gp}(\tilde{t})=\lambda \tilde{t}^{-\alpha}+ \frac{\mu}{1-\tilde{t}}\,,
\label{e.fit}
\end{equation}
with only two adjustable parameters $\lambda$ and $\mu$ once $\alpha=0.85$ is fixed. The remarkable collapse of the whole data set to such a simple equation allows us to interpret the secondary creep regime as a mere crossover from creep to crack growth.

Let us now summarize and discuss the most prominent results of this Letter. First we have shown that casein gels display a remarkable failure scenario similar to that of brittle solids and characterized by the same three successive creep regimes. Here, the primary creep and its power-law exponent [$\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$] are fully accounted for by linear viscoelasticity and local measurements consistently display a homogeneous bulk deformation. Such a link between creep and viscoelasticity is shared by other biopolymer gels with power-law rheology \cite{Gobeaux:2010,Jaishankar:2013} but contrasts with the primary creep of solid materials where smaller exponents are found (0.3 $\lesssim \alpha \lesssim 0.7$). Besides primary creep in solids is irreversible and has been attributed to dislocation interactions in the case of crystalline materials \cite{Miguel:2002} or to localized plastic rearrangements in amorphous materials \cite{Rosti:2010}. Protein gels rather appear to remain purely elastic and reversible throughout primary creep, a hallmark of brittle materials.

The logarithmic fracture growth in the tertiary creep regime constitutes our second important result. Such an evolution is also commonly reported in disordered solid materials displaying brittle rupture and interpreted in the framework of Griffith-like models based on global or local energy barriers \cite{Vanel:2009}. However these approaches all predict exponential scalings for $\tau_f(\sigma)$ while our data are best fitted by the decreasing power-law $\tau_f \sim \sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ [inset of Fig.~\ref{fig1}]. This last key result suggests that thermally activated crack growth is not relevant to the present soft system. Rather the power-law scaling is strikingly reminiscent of the Basquin law of fatigue found for a variety of heterogeneous or cellular materials under cyclic deformation \cite{Kun:2007,Basquin:1910,Suresh:2006}. Basquin law has also been recently predicted by FBMs that combine elastic fibers with a local yield strain and take into account damage accumulation \cite{Kun:2007,Halasz:2012}. Interestingly assuming damage accumulation to be proportional to $\sigma^\gamma$ directly leads to Basquin law with $\beta=\gamma$ and large values of $\gamma$, typically larger than 5 as found here for $\beta$, lead to macroscopic cracks due to the simultaneous rupture of a large number of fibers \cite{Halasz:2012}. More generally FBMs under compressive or elongational load predict three successive creep regimes exactly alike Fig.~\ref{fig2} with similar proportionality between $\tau_{\rm min}$ and $\tau_f$ and finite-time singularity \cite{Nechad:2005,Jagla:2011}. This highlights the relevance of FBMs in the context of creep in protein gels, whose microstructure indeed appears to be formed of strands \cite{Kalab:1983,Roefs:1990}, but also urges to study FBMs in shear geometries to check whether they would be able to predict fracture growth as observed here.

To conclude, the present time- and space-resolved study shows that protein gels qualify as model, brittle soft solids. Prompted by the remarkable simplicity of Eq.~(\ref{e.fit}) that encompasses Andrade creep, the Monkman-Grant relation and finite-time singularity within a single equation, future modeling will undoubtedly focus on the microscopic ingredients that may give rise to this expression. The next experimental step consists in a statistical study of the fluctuations associated to crack nucleation and growth as well as a systematic investigation of other systems in order to check for universality in the irreversible creep rupture of soft solids in general and of biogels such as actin, alginate or agar gels in particular. Such a study is expected to have important implications in understanding the behavior of biomaterials under extreme stress conditions.

@@

\bibitem{Mezzenga:2005} R. Mezzenga, P. Schurtenberger, A. Burbidge and M. Michel, Nature Mater.  {\bf 4}, 729--740 (2005).

@@

\bibitem{Baumberger:2006} T. Baumberger, C. Caroli and D. Martina, Nature Mater. {\bf 5}, 552-555 (2006).

@@

%\bibitem{1990} P.G. Higgs and S.B. Ross-Murphy, Int. J. Biol. Macromol. {\bf 12}, 233--240 (1990).

@@

\bibitem{Tschoegl:1989}  N.W. Tschoegl, {\it The Phenomenological Theory of Linear Viscoelastic Behavior} (Springer, Berlin, Germany, 1989).

@@

\bibitem{Sundararajan:1989} G. Sundararajan, Mater. Sci. Eng. A {\bf 112}, 205-214 (1989).

@@

Supplemental Movie~1 shows the failure of a 4\%~wt. casein gel acidified with 1\%~wt. GDL in a plate-plate geometry for an imposed shear stress $\sigma=120$~Pa. The plate diameter is 50~mm and the gap width is 1~mm. Fractures grow parallel to the vorticity direction, i.e. along the radial direction in this case. The global rheological response is fully similar to that measured in the Couette geometry.
